# COMP1810 - Data Analysis Project

This is the sample project for Data Analysis. We will study:

-   how to create Rproject

-   using Rmd

-   structure the folders and files

-   Step-by-step data analysis

```{r}
# Example block code
#df <- penguins
#summary(df)
```

## Step 0: Loading packages

```{r}
install.packages("here")
```

```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(Hmisc)
```

## Step 1: Read & Understand Data

To understand the data, we have to answer some questions:

1.  What is the shape of dataset? How many rows, how many cols?

    ```{r}
    df <- read_csv("../data/data.csv", show_col_types = FALSE)
    #dim(df)
    #nrow(df)
    #ncol(df)
    cat("The dimension of dataset: number of columns", ncol(df), "number of rows: ", nrow((df)))
    ```

2.  What variables are categorical, numerical, datetime?

    To check the variable with double or numeric type is a categorical column or not, we need to check the distinct (unique) values

    ```{r}
    str(df) # Structure of dataset
    ```

    ```{r}
    describe(df) # provide description for each column
    ```

    ```{r}
    df %>% count(y)
    ```

    **Conclusion**: This dataset has 3 numerical variables `x_1, x_2, x_3`, 1 categorical variable `y`(because it has 2 values only `0` and `1`).

3.  Do we have any issues in this dataset?

    missing values?

    inconsistent types, values?

    duplicate values?

    ```{r}
    names(df) # list all columns of dataset
    colnames(df)
    ```

    ```{r}
    numerical_cols <- names(df)[sapply(df, is.numeric)]
    numerical_cols
    ```

    ```{r}
    # check missing values
    colSums(is.na(df))
    ```

    Conclusion: There is no error (issue) in this dataset.

    ```{r}
    install.packages("skimr")
    library(skimr)
    ```

## Step 2: Explore Dataset - EDA

-   Univariate analysis

    ```{r}
    num_cols <- c("x", "x2", "x3")
    cat_cols <- y
    ```

    ```{r}
    for (col in num_cols){
      vec <- df[[col]]
      print(vec)
      print(class(vec))
      print('=================')
    }
    ```

    ```{r}
    plot_and_save_numeric_cols <- function(df, outdir="../outputs/univariate"){
      if (!dir.exists(outdir)){
        dir.create(outdir, recursive=TRUE)
      }
      for (col in names(df)){
        if (is.numeric(df[[col]])){
          file_path <- file.path(outdir, paste0(col, "_hist_box.png"))
          print(file_path)
          png(file_path, width=1000, height=500)
          par(mfrow=c(1,2))
          hist(df[[col]], main=paste0("Histogram of ", col), col='blue', border='black')
          boxplot(df[[col]], main=paste0("Boxplot of ", col), horizontal = TRUE, col = 'pink')
          dev.off()
        }
      }
      print("All done!!!")
    }

    ```

    ```{r}
    plot_and_save_numeric_cols(df)
    ```

# Example 2: Using Penguins Dataset

```         
penguins data
```

## Step 1: read & understand data

1.  What's the shape of dataset? How many rows, how many cols?

```{r}
df_pen <- penguins
# dim(df_pen)
 cat("The dimension of dataset: number of columns", ncol(df_pen), "number of rows: ", nrow((df_pen)))
```

2.  What variables are categorical, numerical, datetime? To check the variable with double or numeric type is a categorical column or not, we need to check the distinct (unique) values

```{r}
str(df_pen) # Structure of dataset
```

```{r}
describe(df_pen) # provide description for each column
```

```{r}
df_pen %>% count(species)
```

**Conclusion**: This dataset has 5 numerical variables `Bill_len, Bill_dep, flipper_len, boddy_mas, year`, 3 categorical variable `Species, Island, Sex`(because they have unique value to be classified, 2 distinct value for sex, 3 disntict value for species and island).

```{r}
df_pen %>% count(species)
df_pen %>% count(island)
df_pen %>% count(sex)
```

3.  Do we have any issues in this dataset?

    missing values?

    inconsistent types, values?

    duplicate values?

```{r}
names(df_pen) # list all columns of dataset
# colnames(df_pen)
```

```{r}
colSums(is.na(df_pen)) # check missing values
```

```{r}
sapply(penguins, class) # Check data type
```

```{r}
any(duplicated(df_pen)) # check duplicate
cat("The number of dublicated data in penguins: ", sum(duplicated(df_pen)))
```

Conclusion: beside having missing datas (`11 sex, 2 bill_len, 2 bill_dep, 2 fipper_len, 2 body_mas`).

1.  The variable Cateorical variable: Numerical Variable:

2.  The issues

```{r}
# Select penguins with missing value
penguins_with_na <- df_pen %>% 
  filter(if_any(everything(), is.na))
# View result
penguins_with_na
```

```{r}
# find which have more
penguins %>%
  filter(!is.na(sex)) %>%
  count(sex) %>%
  mutate(percent = round(n / sum(n) * 100, 2))
```

### Cleaning the data

```{r}
df_pen$year <- as.Date(paste0(df_pen$year, "-01-01"))
#df_pen$year <- as.POSIXct(paste0(df$year, "-01-01"), format = "%Y-%m-%d")
```

```{r}
str(df_pen$year)
# hoặc
head(df_pen$year)
```

```{r}
colSums(is.na(df_pen))
```

```{r}
df_pen_adeli_torgesen <- df_pen %>% filter(species=="Adelie", island=="Torgersen")
df_pen_adeli_torgesen
```

```{r}
df_pen_adeli_torgersen <- df_pen %>%
  filter(species == "Adelie", island == "Torgersen") %>%
  summarise(
    bill_len_mean = mean(bill_len, na.rm = TRUE),
    bill_len_median = median(bill_len, na.rm = TRUE)
  )

df_pen_adeli_torgersen
```

```{r}
bill_len_mean <- 38.95
df_pen_adeli_torgersen[is.na(df_pen_adeli_torgersen$bill_len), "bill_len"] <- bill_len_mean
bill_len_mean
```

mutate function: create, modify columns

-   for example: with numerical variables, choose to fill missing values by using mean

```{r}


df_pen <- df_pen %>%
  group_by(species, island) %>%
  mutate(
    bill_len_mean = mean(bill_len, na.rm = TRUE),
    bill_len_median = median(bill_len, na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r}
#fill missing value using mutate() and if else()

df_pen <- df_pen %>%
  mutate(
    bill_dep_clean = ifelse(is.na(bill_dep), 
                            mean(bill_dep, na.rm = TRUE), 
                            bill_dep)
  )
```

```{r}
num_cols <- names(df_pen)[sapply(df_pen, is.numeric)]

df_pen[num_cols] <- df_pen[num_cols] %>%
  lapply(function(x) ifelse(is.na(x), median(x, na.rm = TRUE), x)) %>%
  as.data.frame()

df_pen
```

## Step 2: EDA

```{r}
plot_and_save_numeric_cols_for_penguins <- function(df, outdir="../outputs/univariate"){
  if (!dir.exists(outdir)){
    dir.create(outdir, recursive=TRUE)
  }
  for (col in names(df)){
    if (is.numeric(df[[col]])){
      file_path <- file.path(outdir, paste0(col, "_hist_box.png"))
      print(file_path)
      png(file_path, width=1000, height=500)
      par(mfrow=c(1,2))
      hist(df[[col]], main=paste0("Histogram of ", col),xlab = col, col='blue', border='black')
      boxplot(df[[col]], main=paste0("Boxplot of ", col), xlab = col, horizontal = TRUE, col = 'pink')
      dev.off()
    }
  }
  print("All done!!!")
}

```

```{r}
plot_and_save_numeric_cols_for_penguins(df_pen)
```

## Step 3: Pre-process
