# COMP1810 - Practicing data analysis

## Step 0: Loading packages

```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(Hmisc)
```

## Step 1: Read & Understand Data

3 Question to understand about data:

1.  What's the shape of the data? The dimension

```{r}
# data() # check all the dataset
df <- ChickWeight # get the dataset and asign it into an variable?
cat("The dimension of dataset: number of columns", ncol(df), "number of rows: ", nrow((df)))
```

2.  What variables are categorical, numerical, datetime?

```{r}
str(df) # Structure of dataset
```

```{r}
describe(df) # provide description for each column
```

```{r}
sapply(df, class)
```

**Conclusion**: This dataset have 2 numerical variables `weight, time`, and 2 categorical variables `Chick` (with an ordered factor with 50 levels) and `diet` (with factor with 4 levels). And there's no datetime variables in this dataset because the Time is in numerical data type (in days which in reasonable enough as its the count for days).

3.  Do we have any issues in this dataset?

    missing values?

    inconsistent types, values?

    duplicate values?

```{r}
    names(df) # list all columns of dataset
    # colnames(df)
```

```{{r}}
# check missing values
colSums(is.na(df))
```

```{r}
# check duplicated value
any(duplicated(df))
```

```{r}
library(skimr)
skim(df)
```

**Conclusion:** This dataset have no issues, it's clean.

## Step 2: EDA

Univariate analysis

### Numerical Data

analyze numerical data by Selecting them, Getting summary statistics, Plotting their distributions.
We analyze the numerical columns `weight` and `Time` in the dataset. Summary statistics show:

```{r}
num_df <- select_if(df, is.numeric)
summary(num_df) # get statistic of numerical variable in the dataset
```

Distribution of them:

#### Weigth Distribution

```{r}
# Histogram for weight
ggplot(df, aes(x = weight)) + # Set which column goes on the x-axis
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black") + # Add the actual graph type (bars, lines, etc.)
  labs(title = "Distribution of Chick Weights")

ggplot(df, aes(x = weight, fill = Diet)) +
  geom_histogram(binwidth = 10, alpha = 0.6, position = "identity")

ggplot(df, aes(x = Diet, y = weight)) +
  geom_boxplot(fill = "lightblue") + # Boxplot is better for grouped comparation
  labs(title = "Weight Distribution by Diet")

ggplot(df, aes(x = weight)) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  labs(title = "Density of Chick Weights")
```

The weight distribution is right-skewed, meaning most chicks have lower weights but a few have high weights. Using color grouping `(fill = Diet)`, we can compare how the weight or time distribution varies across different diets. This allows us to visually assess whether certain diets lead to higher or lower chick weights.

#### Time Distribution

```{r}
# Histogram for Time
ggplot(df, aes(x = Time)) +
  geom_histogram(binwidth = 2, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Time (Days)")

ggplot(df, aes(x = Time, fill = Diet)) +
  geom_histogram(binwidth = 10, alpha = 0.6, position = "identity")

ggplot(df, aes(x = Time)) +
  geom_density(fill = "lightgreen", alpha = 0.5) +
  labs(title = "Density of Chick Time")
```

The time distribution is uniform (or slightly right-skewed), since chicks were weighed at regular intervals over time.

**Conclusion:** Visualizing `weight` and `Time` by `Diet` shows how different diets may influence growth patterns.

### Categorical Data

For categorical variables, EDA usually includes:

-   Count of unique values (table())

-   Check for imbalance (e.g., are some categories more common?)

-   Visualize with bar plots

-   Interpret patterns (e.g., are diets balanced?)

We analyze the categorical columns `Chick` and `Diet` in the dataset.

#### The frequency of each unique values:

```{r}
table(df$Chick)
table(df$Diet)
```

#### Bar Plot for Diet

Each chick is measured multiple times, so Chick will also appear multiple times — the Diet bar chart shows ***number of observations***, not number of unique chicks.

```{r}
ggplot(df, aes(x = Diet)) +
  geom_bar(fill = "coral", color = "black") +
  labs(title = "Count of Observations per Diet",
       x = "Diet Type", y = "Number of Observations")

ggplot(df, aes(x = weight)) +
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black") +
  facet_wrap(~Diet) # For separate plots per Diet
```

To see how many chick was feeded with each type of diet:

```{r}
# Remove duplicates: one row per Chick
chick_diet <- df %>%
  select(Chick, Diet) %>% # Select only Chick and Diet column in dataset
  distinct() # Remove duplicate value

# Count how many chicks in each Diet
chick_diet_summary <- chick_diet %>%
  group_by(Diet) %>% # group by diet (diet 1 2 3 4)
  summarise(n_chicks = n()) # Count how many chicks are in each diet → n() counts rows per group.

# show total of chicks per diet with diagram
ggplot(chick_diet_summary, aes(x = Diet, y = n_chicks)) + 
  geom_col(fill = "seagreen", color = "black") +
 labs(title = "Number of Unique Chicks per Diet",
       x = "Diet Type", y = "Number of Chicks")
```

To check whether if the groups are balance (which can be seen by the diagram above is not)

```{r}
chick_diet_summary %>%
  mutate(percent = round(n_chicks / sum(n_chicks) * 100, 1))
```

**Conclusion:**

-   There's `50 chicks` in the dataset, they are assigned to `1 of the 4 diet group` and being observed over multiple days

-   The diet groups is not balanced: The first diet group have twice as each of the rest diet, tho those diet group is balanced with each other (10 chick per diet)

### Bivariate analysis

To study about the relationship between two variables

#### Mean weight per diet

To show which diet give the most value in inceasing weight of the chicks

```{r}
df %>%
  group_by(Chick) %>%
  filter(Time == max(Time)) %>%     # Get the last weight of each Chick
  ungroup() %>%
  group_by(Diet) %>%
  summarise(mean_final_weight = mean(weight))
```

#### Weight over Time per Diet

This observation will show which diet increase the weight of the chick most quickly

```{r}
ggplot(df, aes(x = Time, y = weight, color = Diet)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  labs(title = "Chick Weight Over Time by Diet", x = "Time (days)", y = "Weight (grams)")

# df %>%
#  group_by(Diet, Time) %>%
#  summarise(mean_weight = mean(weight), .groups = "drop") %>%
#  ggplot(aes(x = Time, y = mean_weight, color = Diet)) +
#  geom_line(size = 1.2) +
#  geom_point() +
#  labs(title = "Mean Chick Weight Over Time per Diet", x = "Time (days)", y = "Mean Weight")
```